\documentclass{article}
\usepackage{iclr2026_conference,times}
\input{math_commands.tex}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs}
\input{jay_color_utils}

% our new packages--alphabetial order
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[table]{xcolor}
\algtext*{EndFor}
\algrenewcommand\algorithmicdo{}
\algtext*{EndIf}
\algrenewcommand\algorithmicthen{}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bbm}
\usepackage{booktabs}
\usepackage{bm}
\usepackage{caption}
\captionsetup[figure]{name=Figure, labelfont=bf, labelsep=colon} 
\usepackage{duckuments}
\usepackage{enumitem}
\usepackage{etoolbox}
\usepackage{float}
\usepackage{graphicx}
\usepackage{lipsum}
\usepackage{listings}
\usepackage{mdframed}
\usepackage{multirow}
\usepackage{soul}
\usepackage{stmaryrd}
\usepackage{subcaption}
\usepackage{trimclip}
\usepackage{url}
\usepackage{xcolor}
\usepackage{xspace}
\usepackage[normalem]{ulem}
\usepackage{hyperref}
\usepackage[most]{tcolorbox}

% author list utilities: made by Jaeyeon
\makeatletter
\def\xthanks#1{\protected@xdef\@thanks{\@thanks\protect\footnotetext{#1}}}
\makeatother
\newcommand{\xsup}[1]{\rlap{\textsuperscript{\normalfont#1}}}
\newtheorem{definition}{Definition}
\newtheorem{prop}{Proposition}
\newtheorem{proposition}{Proposition}
\newtheorem*{corollary*}{Corollary} % disable the numbering
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
\newcommand{\mask}{\mathbf{m}}
\newcommand{\rebuttal}[1]{\textcolor{black}{#1}}
\def\rvx{{\mathbf{x}}}
\def\rvy{{\mathbf{y}}}
\def\rvz{{\mathbf{z}}}
\def\rmd{{\mathrm{d}}}


\title{Fine-Tuning Masked Diffusion for Provable\\ Self-Correction}


\author{%
\textbf{
\hspace{-17.1em}
Jaeyeon Kim}$^{1\star}$\xthanks{\hspace{-1.0em}$^\star$ Jaeyeon Kim and Seunggeun Kim contributed equally; 
Taekyun Lee is also a co–first author.} \quad\hspace{0.1em}
\textbf{Seunggeun Kim}$^{2\star}$ \quad\hspace{0.1em}
\textbf{Taekyun Lee}$^{2\star}$ \\
\hspace{-16.9em}
\textbf{David Z. Pan}$^{2}$ \quad\hspace{0.1em}
\textbf{Hyeji Kim}$^{2}$ \quad\hspace{0.1em}
\textbf{Sham Kakade}$^{1,3}$ \quad\hspace{0.1em}
\textbf{Sitan Chen}$^{1}$
\\[0.5ex]
\hspace{-17.2em}
~$^1$Harvard University \quad
$^2$UT Austin \quad
$^3$Kempner Institute
\vspace{-0.15in}
} 

\newcommand{\PRISM}{{PRISM }}
\iclrfinalcopy
\begin{document}
\maketitle
\begin{abstract}
\vspace{-0.1in}
A natural desideratum for generative models is \emph{self-correction}--detecting and revising low-quality tokens at inference. While Masked Diffusion Models (MDMs) have emerged as a promising approach for generative modeling in discrete spaces, their capacity for self-correction remains poorly understood. Prior attempts to incorporate self-correction into MDMs either require overhauling MDM architectures/training or rely on imprecise proxies for token quality, limiting their applicability. Motivated by this, we introduce PRISM--{\bf P}lug-in {\bf R}emasking for {\bf I}nference-time {\bf S}elf-correction of {\bf M}asked Diffusions--a lightweight, model-agnostic approach that applies to any pretrained MDM. Theoretically, PRISM defines a self-correction loss that provably learns per-token quality scores, without RL or a verifier. These quality scores are computed in the same forward pass with MDM and used to detect low-quality tokens. Empirically, PRISM advances MDM inference across domains and scales: Sudoku; unconditional text (170M); and code with LLaDA (8B).
\vspace{-0.15in}
\end{abstract}
\input{1_Introduction}
\input{2_Prelim}
\input{3_Method}
\input{4_Experiments}
\input{5_Conclusion}
\newpage
\bibliography{main}
\bibliographystyle{iclr2026_conference}
\newpage
\appendix
\tableofcontents
\input{Appendix_related_works}
\input{Appendix_theory}
\input{Appendix_pseudocodes}
\input{Appendix_exp_details}
\end{document}
