\section{Discussion}
Much of the current tooling around pre-training, post-training, and inference for text generation has been built around a key modeling choice: next-token prediction training. \glspl{mdlm} are an expressive class of models trained to in-fill masked sequences, requiring additional training and inference compute. In our work, we find that this additional compute has many uses beyond just accelerating inference and warrants \textit{rethinking} how these models are utilized. For instance, the ability to in-fill unlocks new prompting techniques, like the proposed reasoning-as-infilling framework, along with new data generation and post-training methods. Large-scale language models trained on next-token prediction have revolutionized text generation. Our work provides evidence that models such \glspl{mdlm}, trained on alternative objectives, unlock new capabilities not readily available to \gls{ntp} approaches.



















 












\subsection{Limitations}
State-of-the-art \gls{ntp} language models achieve significant benefits from long reasoning chains \citep{guo2025deepseek, openai2025o3}. While the \gls{mdlm} framework introduces new possibilities, \glspl{mdlm} come with computational costs that make the current available models impractical for long-context tasks. Most notably, for a sequence of length $\ell$, an \gls{ntp} model needs to make $\ell$ predictions, while an \gls{mdlm} model makes $\mathcal{O}(\ell^2)$ predictions. Moreover, current \glspl{mdlm} do not naturally support inference-optimizations, such as caching, for any-order decoding. We are hopeful that our findings, which highlight the benefits and sampling behavior of current models, can help guide architectural decisions for future models that address these limitations. 

\section{Acknowledgments}
The authors would like to thank Subham Sahoo, Rico Angell, Chandhru Karthick, Mark Goldstein, Shenglong Wang, and Dinghuai Zhang for their valuable feedback. 

This work was partly supported by the NIH/NHLBI Award R01HL148248, NSF Award 1922658 NRT-HDR:
FUTURE Foundations, Translation, and Responsibility for Data Science, NSF CAREER Award 2145542,
ONR N00014-23-1-2634, and Apple. Additional support was provided by a Fellowship from the Columbia
Center of AI Technology. This work was also supported by IITP with a grant funded by the MSIT of the Republic of Korea in connection with the Global AI Frontier Lab International Collaborative Research.










