\documentclass{article}


% if you need to pass options to natbib, use, e.g.:
\PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2025


% ready for submission
% \usepackage{neurips_2025}


% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
\usepackage[preprint]{neurips_2025}


% to compile a camera-ready version, add the [final] option, e.g.:
%     \usepackage[final]{neurips_2025}


% to avoid loading the natbib package, add option nonatbib:
% \usepackage[nonatbib]{neurips_2025}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
% \usepackage{xcolor}         % colors
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{cleveref}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{multirow} % For \multirow
\usepackage{siunitx}  % For S column type (decimal alignment) and formatting numbers
\usepackage[svgnames]{xcolor} 
\usepackage{tcolorbox}
\usepackage{wrapfig}
\tcbuselibrary{skins}         
\usepackage[all=normal, paragraphs=tight,floats=tight]{savetrees}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}



\title{Any-Order GPT as Masked Diffusion Model:\\Decoupling Formulation and Architecture}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.
\makeatletter
\renewcommand{\@fnsymbol}[1]{\ensuremath{\ifcase#1\or \dagger\or \ddagger\or \mathsection\or \mathparagraph\or \|\or **\or \dagger\dagger \or \ddagger\ddagger \else\@ctrerr\fi}}
\makeatother

\author{
Shuchen Xue$^{1,2}$\thanks{Email: \texttt{xueshuchen17@mails.ucas.edu.cn}}
, Tianyu Xie$^3$, Tianyang Hu$^4$\thanks{Corresponding Authors}, Zijin Feng$^5$\\
\textbf{Jiacheng Sun$^5$$^\ddagger$, Kenji Kawaguchi$^4$, Zhenguo Li$^5$, Zhi-Ming Ma$^{1,2}$}\\
$^{1}$ University of Chinese Academy of Sciences\\
$^{2}$ Academy of Mathematics and Systems Science, Chinese Academy of Sciences\\
$^{3}$ School of Mathematical Sciences, Peking University\\
$^{4}$ National University of Singapore\\
$^{5}$ Huawei Noah’s Ark Lab
}

\begin{document}



\maketitle


\begin{abstract}
Large language models (LLMs) predominantly use autoregressive (AR) approaches, but masked diffusion models (MDMs) are emerging as viable alternatives. A key challenge in comparing AR and MDM paradigms is their typical architectural difference: AR models are often decoder-only, while MDMs have largely been encoder-only. This practice of changing both the modeling paradigm and architecture simultaneously makes direct comparisons unfair, as it's hard to distinguish whether observed differences stem from the paradigm itself or the architectural shift. This research evaluates MDMs within a decoder-only framework to: (1) equitably compare MDM (as Any-Order AR, or AO-AR) and standard AR paradigms. Our investigation suggests that the standard AO-AR objective, which averages over all token permutations, may benefit from refinement, as many permutations appear less informative compared to the language's inherent left-to-right structure. (2) Investigate architectural influences (decoder-only vs. encoder-only) within MDMs. We demonstrate that while encoder-only MDMs model a simpler conditional probability space, decoder-only MDMs can achieve dramatic generation speedups ($\sim25\times$) and comparable perplexity with temperature annealing despite modeling a vastly larger space, highlighting key trade-offs. This work thus decouples core paradigm differences from architectural influences, offering insights for future model design. Code is available at \url{https://github.com/scxue/AO-GPT-MDM}.
\end{abstract}

\input{secs/intro}
\input{secs/prelim}
\input{secs/results}
\input{secs/methods}
\input{secs/relate}
\input{secs/discussion}




















% \begin{ack}
% Use unnumbered first level headings for the acknowledgments. All acknowledgments
% go at the end of the paper before the list of references. Moreover, you are required to declare
% funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
% More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2025/PaperInformation/FundingDisclosure}.


% Do {\bf not} include this section in the anonymized submission, only in the final paper. You can use the \texttt{ack} environment provided in the style file to automatically hide this section in the anonymized submission.
% \end{ack}

% \bibliographystyle{unsrtnat}
\bibliographystyle{ieeetr}
\bibliography{refs.bib}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix
\newpage
\input{secs/app}



\end{document}