\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsfonts, amsthm, bm}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{xcolor}
\usepackage{algorithm}
\IfFileExists{algpseudocode.sty}{
  \usepackage{algpseudocode}
}{
  \newenvironment{algorithmic}[1][]{
    \begin{enumerate}
    \setlength{\itemsep}{1pt}
    \setlength{\parskip}{0pt}
  }{
    \end{enumerate}
  }
  \newcommand{\State}{\item}
  \newcommand{\For}[1]{\item \textbf{for} ##1 \textbf{do}}
  \newcommand{\EndFor}{\item \textbf{end for}}
  \newcommand{\If}[1]{\item \textbf{if} ##1 \textbf{then}}
  \newcommand{\EndIf}{\item \textbf{end if}}
  \newcommand{\Comment}[1]{\quad{\footnotesize\textit{##1}}}
}
\usepackage{enumitem}
\usepackage{caption}
\IfFileExists{cleveref.sty}{
  \usepackage[capitalize,noabbrev]{cleveref}
}{
  \newcommand{\cref}[1]{\ref{##1}}
  \newcommand{\Cref}[1]{\ref{##1}}
}

% Commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\KL}{\mathrm{KL}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\cM}{\mathcal{M}}
\newcommand{\cK}{\mathcal{K}}
\newcommand{\cU}{\mathcal{U}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cV}{\mathcal{V}}
\newcommand{\cS}{\mathcal{S}}
\newcommand{\cA}{\mathcal{A}}
\newcommand{\cQ}{\mathcal{Q}}
\newcommand{\cC}{\mathcal{C}}
\newcommand{\cI}{\mathcal{I}}
\newcommand{\cL}{\mathcal{L}}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bphi}{\bm{\phi}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bc}{\mathbf{c}}
\newcommand{\bh}{\mathbf{h}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bfm}{\mathbf{m}}
\newcommand{\bp}{\mathbf{p}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\mask}{\ensuremath{\mathtt{[M]}}}

\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}

\title{Any-Order Adaptive Editing for Masked Diffusion LLMs}
\author{}
\date{}

\begin{document}
\maketitle

% ======================================================================
\begin{abstract}
% ======================================================================
Masked diffusion language models (MDLMs) generate text by iteratively unmasking tokens in parallel, but most inference methods still use fixed confidence thresholds or block schedules.
We propose \textbf{Any-Order Adaptive Editing (AOAE)}, a lossy self-speculative framework for MoE MDLMs.
A fast \emph{hard-routed auxiliary} runs in an aggressive low-cost mode---for example S-mode or a heuristic cache/freezing policy---while a slower verifier runs the normal higher-quality inference mode.
The verifier accepts easy auxiliary drafts, rejects or remasks stale regions, and can reset the auxiliary cache when disagreement grows.
AOAE separates transient speculative acceptance from persistent stability caching, and optionally learns a cache/access policy with GRPO.
This yields a tunable speed--quality frontier: conservative settings recover verifier-quality decoding, while more aggressive settings trade quality for speed through controlled verifier rejection.
In implementation, the paper evaluation anchors the frontier with verifier-quality blockwise baselines and sweeps AOAE with named operating points over verifier cadence, agreement threshold, unmask budget, remasking, the one-step $\cK_{\text{spec}}$ skip, and policy temperature rather than by changing policy temperature alone; prompt formatting uses chat-template auto-fallback plus a GSM8K answer-format scaffold in math mode, non-chat prompts include special tokens, and confidence-style baseline rows are scored on fully resolved (mask-completed) outputs.
\end{abstract}

% ======================================================================
\section{Introduction}
\label{sec:intro}
% ======================================================================

Masked diffusion language models (MDLMs) generate text by reversing a token-masking process, producing all positions in parallel rather than left-to-right~\cite{austin2021structured,sahoo2024simple,shi2024simplified,lou2024discrete}.
This enables \emph{any-order} generation: the model can resolve dependencies in whatever sequence best suits the problem.
Recent scaling has produced frontier-class MDLMs---LLaDA~2.0 at 100B parameters~\cite{nie2025llada2} and LLaDA~2.1 with Token-to-Token (T2T) editing for retroactive error correction~\cite{llada21}---but inference remains governed by static heuristics.
LLaDA~2.1 uses fixed confidence thresholds $\tau_{\text{mask}}$ and $\tau_{\text{edit}}$ and decodes in rigid semi-autoregressive blocks; Fast-dLLM~\cite{wu2025fastdllm} applies a single confidence threshold for variable-rate unmasking.
These approaches require per-task tuning, degrade on long sequences~\cite{jazbec2025learning}, and impose uniform compute regardless of position difficulty.

Three lines of work have begun to address these limitations. Kim et al.~\cite{kim2025trainworst} showed that adaptive token ordering boosts MDM accuracy on Sudoku from $<$7\% to ${\approx}$90\%, and Jazbec et al.~\cite{jazbec2025learning} trained a lightweight RL policy to decide \emph{which} tokens to unmask, outperforming heuristic samplers---but their formulation covers only unmasking, ignoring self-correction and caching.
PRISM~\cite{kim2025prism} provides model-agnostic self-correction via remasking, but uses a fixed threshold rather than a learned policy.
dKV-Cache~\cite{ma2025dkv} and Elastic-Cache~\cite{elastic_cache} accelerate inference by caching stable KV states, but they \emph{constrain} the decoding trajectory to suit the cache (block-based or prefix-restricted unmasking), sacrificing diffusion's any-order flexibility.

AOAE is built on a simple observation: MoE MDLMs expose a natural drafter--verifier split.
With standard hard top-$k$ routing, LLaDA~2.1-mini activates ${\sim}$1.4B of 16B parameters per token; the same frozen model can also be run in a higher-quality verifier mode.
To make the speed--quality tradeoff nontrivial, AOAE does \emph{not} rely only on hard-vs.-soft routing.
The hard-routed auxiliary also runs an aggressive inference shortcut (e.g., S-mode, dKV/Elastic-style caching, or a learned cache policy), while the verifier runs normal inference and periodically checks the drafted state.
If verifier disagreement grows, AOAE rejects stale regions, remasks uncertain tokens, and resets the auxiliary cache.
This turns verifier rejection into a cache-realignment mechanism and makes the learned stability cache an additional contribution rather than a prerequisite for speculation.

% ======================================================================
\section{Background and Notation}
\label{sec:background}
% ======================================================================

\subsection{Masked Diffusion Models}
\label{sec:bg_mdm}

Let $\bx_0 = (x_0^1, \dots, x_0^L) \in \cV^L$ be a token sequence with vocabulary $\cV$ and mask token $\mask$.
The forward process independently corrupts each token:
\begin{equation}
    q(x_t^k \mid x_0^k) = \begin{cases}
        1 - t, & \text{if } x_t^k = x_0^k, \\
        t, & \text{if } x_t^k = \mask, \\
        0, & \text{otherwise,}
    \end{cases}
    \label{eq:forward}
\end{equation}
for $t \in [0, 1]$, so at $t{=}1$ all tokens are masked.
A denoising network $p_{\btheta}$ is trained via the ELBO~\cite{ou2024your,sahoo2024simple}:
\begin{equation}
    \cL(\btheta) = \E_{t,\, \bx_0,\, \bx_t} \left[ \frac{1}{t} \sum_{k=1}^L \ind[x_t^k = \mask] \log p_{\btheta}(x_0^k \mid \bx_t) \right].
    \label{eq:elbo}
\end{equation}

At inference, the model starts from $\by_T = (\mask, \dots, \mask)$ and at each step $t$ computes per-position distributions $p_t^k := p_{\btheta}(y_0^k {=} \cdot \mid \by_t)$.
A sampling strategy selects a subset $\cU_t \subseteq \cM_t$ of the masked positions $\cM_t := \{k : y_t^k = \mask\}$ to unmask:
\begin{equation}
    y_{t-1}^k = \begin{cases}
        y \sim p_t^k, & \text{if } k \in \cU_t, \\
        y_t^k, & \text{otherwise.}
    \end{cases}
    \label{eq:sampling}
\end{equation}
The choice of $\cU_t$ critically affects both speed and quality~\cite{jazbec2025learning,wu2025fastdllm}.

\subsection{LLaDA 2.1: Draft-and-Edit Decoding}
\label{sec:bg_llada}

LLaDA~2.1~\cite{llada21} extends unmasking with T2T editing via confidence thresholds:
\begin{align}
    \Gamma_t &= \bigl\{ k \in \cM_t \;\big|\; \max_v p_{\btheta}(v \mid \by_t) > \tau_{\text{mask}} \bigr\} & \text{(M2T unmasking)}, \label{eq:gamma} \\
    \Delta_t &= \bigl\{ k \notin \cM_t \;\big|\; \max_v p_{\btheta}(v \mid \by_t) > \tau_{\text{edit}} \;\land\; \arg\max_v \neq y_t^k \bigr\} & \text{(T2T editing)}. \label{eq:delta}
\end{align}
Two modes are offered: \emph{Speedy (S)} with aggressive thresholds and \emph{Quality (Q)} with conservative ones, but both are fixed hyperparameters.

\subsection{KV-Caching in Diffusion Models}
\label{sec:bg_dkv}

In MDLMs, bidirectional attention causes KV states to evolve across steps, making naive caching fragile.
dKV-Cache~\cite{ma2025dkv} caches high-confidence positions and recomputes only changed ones, achieving $2$--$10\times$ speedup.
The per-token, per-layer \emph{KV drift} quantifies staleness:
\begin{equation}
    \Delta^{k}_{t,\ell}
    \;:=\;
    \left\lVert K^{k}_{t,\ell} - K^{k}_{t-1,\ell} \right\rVert_2
    +
    \left\lVert V^{k}_{t,\ell} - V^{k}_{t-1,\ell} \right\rVert_2.
    \label{eq:kv_drift}
\end{equation}
Drift is empirically small for most steps and grows with layer depth, suggesting that recomputing shallow layers at every step is wasteful.

Elastic-Cache~\cite{elastic_cache} exploits this structure with three mechanisms: (i)~locality via a sliding window $\cM_t^\beta$ over masked positions; (ii)~an attention-derived staleness proxy based on cosine similarity of attention weights to the most-attended decoded token:
\begin{equation}
    \sigma_{t,\ell}
    :=
    \frac{
      \langle S_{t-1,\ell}[\cM_t^\beta, T_{t-1,\ell}],\; S_{t,\ell}[\cM_t^\beta, T_{t-1,\ell}] \rangle
    }{
      \| S_{t-1,\ell}[\cM_t^\beta, T_{t-1,\ell}] \|_2 \;
      \| S_{t,\ell}[\cM_t^\beta, T_{t-1,\ell}] \|_2
    },
    \label{eq:attn_cosine}
\end{equation}
where $T_{t,\ell} := \arg\max_{k \in \cD_{<t}} \sum_{q \in \cM_t^\beta} S_{t,\ell}[q,k]$; and (iii)~depth-selective refresh from a boundary layer $\ell_t^\star := \min\{\ell : \sigma_{t,\ell} < \gamma\}$, recomputing only layers $\ell > \ell_t^\star$.

These methods improve cache efficiency but do so by \emph{constraining} the decoding trajectory---block-based scheduling, prefix-restricted unmasking---trading diffusion's any-order flexibility for a more regular access pattern.

\subsection{Self-Correction via PRISM}
\label{sec:bg_prism}

PRISM~\cite{kim2025prism} trains a lightweight adapter to estimate per-token quality scores $q_t^k$---the probability that a generated token matches the ground truth---via binary cross-entropy on (corrupted, clean) pairs.
Equivalently, PRISM asks whether a token should remain in place under one-step self-consistency: if the score is low, the token is \emph{remasked} (reverted to $\mask$) and regenerated.
In our setting, the same idea can be applied on top of the verifier while scoring an auxiliary-proposed sequence: the verifier-side score becomes a pointwise keep/remask signal, and a PRISM-like head can further be trained to predict verifier disagreement or correction risk directly.
This is strictly simpler than T2T editing: remasking defers to the base model's well-trained M2T denoising rather than learning a replacement distribution conditioned on the current (possibly incorrect) token, preserving the any-order property.

\subsection{Learned Unmasking Policies}
\label{sec:bg_rl}

Jazbec et al.~\cite{jazbec2025learning} formalize MDLM sampling as an MDP where a single-layer transformer policy $\pi_\phi$ outputs per-position Bernoulli unmasking decisions $u_t^k \sim \text{Ber}(\sigma(f_\phi(c_t^k)))$ based on confidences $c_t^k := \max_v p_t^k(v)$, trained via GRPO~\cite{shao2024deepseekmath} with a multiplicative reward:
\begin{equation}
    R(\by, \by_t) = r(\by, \by_t) \cdot \left(1 - \frac{T - t}{T}\right)^\alpha,
    \label{eq:rl_reward}
\end{equation}
where $r(\cdot)$ is task-specific correctness and $\alpha$ controls the speed--accuracy tradeoff.
This outperforms heuristic samplers but addresses only unmasking---no self-correction, caching, or speculative inference.

% ======================================================================
\section{Method: Any-Order Adaptive Editing}
\label{sec:method}
% ======================================================================

AOAE extends the learned-policy framework of \Cref{sec:bg_rl} to a full speculative diffusion architecture, formalized as a finite-horizon MDP $(\cS, \cA, P, R, T)$ over a dual-model system.

\subsection{Dual-Model MoE Architecture}
\label{sec:soft_moe}

LLaDA~2.1-mini is a 16B-parameter MoE model with $E{=}128$ routed experts per layer, of which $k{=}8$ are selected per token via hard top-$k$ routing, so only ${\sim}$1.4B parameters are active per forward pass.
AOAE instantiates two operational modes of the \emph{same} frozen model with different routing and separate cache policies:

\paragraph{Hard-routed auxiliary (fast).}
Standard hard top-$k$ routing (${\sim}$1.4B active).
Serves as the draft model: produces token predictions for all positions and may run with any plug-and-play cache/freezing mechanism $\cC_{\text{aux}}$ (heuristic dKV/Elastic-style rules, learned stable-cache predictions, or other approximate reuse policies).
This path is intentionally faster and lower-quality by construction: the auxiliary uses a known speed-oriented inference mode, and it may take $K_{\text{draft}} > 1$ cheap microsteps between verifier calls.

\paragraph{Soft-routed primary (slow, expressive).}
Replaces hard routing with temperature-controlled soft routing:
\begin{equation}
    w_i^{\text{soft}} = \frac{\exp(z_i / \tau_r)}{\sum_{j=1}^E \exp(z_j / \tau_r)}, \quad i = 1, \dots, E,
    \label{eq:soft_route}
\end{equation}
where $z_i$ are router logits and $\tau_r > 0$ is the routing temperature.
At $\tau_r \to 0$ this recovers hard routing; at $\tau_r > 0$ all experts participate.
Both modes share frozen weights; routing is toggled in-place, avoiding two full model copies.
The primary runs the original high-quality inference mode and acts as verifier/corrector: it scores the drafted state, emits verifier logits, supplies a PRISM-style keep/remask score, and can clear $\cC_{\text{aux}}$ when the fast path drifts too far.
Conceptually, $\cK_{\text{spec}}$ is a short-lived frontier of drafted candidates awaiting validation, while only $\cK_{\text{stable}}$ is a persistent cache.

The operating frontier is controlled by the verifier mode, verifier cadence, agreement threshold, unmask budget, remasking policy, drafter shortcut, $\cK_{\text{spec}}$ skip, draft horizon $K_{\text{draft}}$, policy temperature $\tau_\pi$, and composition strength $\gamma$.
At conservative settings, the verifier dominates and quality is preserved; at aggressive settings, the cached drafter runs farther ahead and the verifier rejects more often.

\subsection{State: Soft-Masked Representations}
\label{sec:state}

The policy's state must convey predictive uncertainty, not just binary mask indicators.
Following Hersche et al.~\cite{hersche2025softmasked}, we construct a soft-masked embedding per position.
Let $p_t^k \in \Delta^{|\cV|-1}$ be the primary's predictive distribution and $c_t^k := \max_v p_t^k(v)$ the confidence.
The soft-masked embedding is:
\begin{equation}
    \bh_t^k = \lambda(p_t^k)\, \mathbf{E}_{\mask} + \bigl(1 - \lambda(p_t^k)\bigr) \sum_{j \in \text{top-}K(p_t^k)} \pi_j^k\, \mathbf{E}_{v_j},
    \label{eq:softmask}
\end{equation}
where $\pi_j^k = p_t^k(v_j) / \sum_{j'} p_t^k(v_{j'})$ are renormalized top-$K$ probabilities and the gating function is:
\begin{equation}
    \lambda(p_t^k) = \omega_s \cdot \sigma\!\bigl(\omega_a(-H(p_t^k) - \omega_b)\bigr),
    \label{eq:gating}
\end{equation}
with entropy $H(p_t^k)$ and learnable parameters $\omega = (\omega_s, \omega_a, \omega_b)$.
At low confidence (high entropy), $\lambda \to \omega_s$ and the mask embedding dominates; at high confidence, $\lambda \to 0$ and the top-$K$ token embeddings dominate.

The full state at verifier event $t$ is $s_t = (\bH_t, \bfm_t, \bm{\rho}_t, \bm{\alpha}_t, t/T)$, where $\bH_t = (\bh_t^1, \dots, \bh_t^L)$, $m_t^k = \ind[y_t^k = \mask]$, $\rho_t^k \in [0,1]$ is the verifier-side PRISM / PRISM-like keep score on the drafted sequence, and $\alpha_t^k = \ind[\rho_t^k \ge \delta_{\text{acc}}]$ is the binary verifier acceptance signal.

The gating parameters $(\omega_s, \omega_a, \omega_b)$ are trained jointly with the policy via GRPO.
To allow gradients to flow through them without re-running the full base model, \texttt{SoftMaskedState.forward()} returns both the embedding $\bh_t^k$ and the intermediate weighted token sum $\mathbf{w}_t^k = \sum_{j \in \text{top-}K} \pi_j^k \mathbf{E}_{v_j}$.
During the GRPO backward pass, $\mathbf{w}_t^k$ (stored in the trajectory) and the detached entropy $H_t^k$ are passed to \texttt{recompute\_h\_t()}, which evaluates $\bh_t^k = \lambda(H_t^k)\,\mathbf{E}_{\mask} + (1-\lambda(H_t^k))\,\mathbf{w}_t^k$ using the current live $\omega$ parameters, allowing autograd to propagate through $\omega$ at ${\sim}$200 MB overhead per rollout group (vs.\ ${\sim}$16 GB if storing full-vocabulary logits).

\subsection{Action: Unmasking, Remasking, and Optional Cache Control}
\label{sec:action}

The system exposes four stochastic policy heads, all conditioned on verifier-side signals:
\begin{align}
    u_t^k &\sim \text{Ber}\bigl(\sigma(f_{\bphi^{\text{unmask}}}(s_t)^k)\bigr), & &\text{(Unmask)}, \label{eq:action_u} \\
    r_t^k &\sim \text{Ber}\bigl(\sigma(f_{\bphi^{\text{remask}}}(s_t)^k)\bigr), & &\text{(Remask / evict)}, \label{eq:action_r} \\
    \kappa_t^k &\sim \text{Ber}\bigl(\sigma(f_{\bphi^{\text{cache}}}(s_t)^k)\bigr), & &\text{(Stable Cache; optional)}, \label{eq:action_c} \\
    z_t^k &\sim \text{Ber}\bigl(\sigma(f_{\bphi^{\text{access}}}(s_t)^k)\bigr), & &\text{(Access / Refresh; optional)}, \label{eq:action_z} \\
    \alpha_t^k &= \ind[\rho_t^k \ge \delta_{\text{acc}}], & &\text{(Verifier accept / reuse signal)}. \label{eq:action_alpha}
\end{align}
Here $\rho_t^k$ is produced by primary--auxiliary agreement, a verifier-side PRISM module, or a PRISM-like correction head.
Thus remasking is learned, but it is not blind: low verifier agreement or low PRISM score makes $r_t^k$ more likely, while high $\alpha_t^k$ makes stable caching and composed prediction safer.

The stochastic policy action $\mathbf{a}_t = (\bu_t, \mathbf{r}_t, \bm{\kappa}_t, \mathbf{z}_t) \in \{0,1\}^{4L}$ has a tractable factorized likelihood:
\begin{equation}
    \pi_{\bphi}(\mathbf{a}_t \mid s_t) = \prod_{k=1}^L \prod_{d \in \{\text{u}, \text{r}, \text{c}, \text{a}\}} \bigl(\sigma(f_{\bphi^d}(s_t)^k)\bigr)^{a_t^{d,k}} \bigl(1 - \sigma(f_{\bphi^d}(s_t)^k)\bigr)^{1 - a_t^{d,k}}.
    \label{eq:policy_likelihood}
\end{equation}

Validity is enforced via logit masking before the sigmoid:
unmask applies only to masked positions ($u_t^k = 0$ if $m_t^k = 0$);
remask applies only to unmasked positions; stable cache and remask are mutually exclusive ($\kappa_t^k \cdot r_t^k = 0$).
The access head $z_t^k$ is unrestricted because it predicts which positions to refresh regardless of mask state.
In conservative ablations, $r_t$ can be forced to a verifier-threshold rule and $\kappa_t$ can be disabled; the full AOAE model learns both.

We choose unmasking and remasking over the T2T editing approach since the latter requires the base model to produce a replacement token conditioned on the current (possibly incorrect) token---learning $p(x_0^k \mid \bx_t, y_t^k \neq \mask)$ on artificially corrupted data.
This is harder than the standard M2T objective $p(x_0^k \mid \bx_t, y_t^k = \mask)$ and breaks the any-order property.
Remasking simply returns a position to $\mask$, after which the base model's well-trained denoising handles it.

\subsection{Inference: Lossy Draft--Validate Loop}
\label{sec:transition}

\begin{algorithm}[!ht]
\caption{AOAE Lossy Draft--Validate Inference}
\label{alg:aoae}
\begin{algorithmic}[1]
\State \textbf{Input:} Drafter $p_{\btheta}^{\text{hard}}$ with plug-and-play cache $\cC_{\text{aux}}$, verifier $p_{\btheta}^{\text{soft}}$, verifier head $h_\psi$ (PRISM or PRISM-like), policy $\pi_{\bphi}$, prompt $\bx$, verifier events $T$, draft horizon $K_{\text{draft}}$
\State \textbf{Init:} $\by_T \leftarrow (\mask, \dots, \mask)$, \; $\cK_{\text{spec}} \leftarrow \emptyset$, \; $\cK_{\text{stable}} \leftarrow \emptyset$, \; $\cC_{\text{aux}} \leftarrow \emptyset$
\For{$t = T, T{-}1, \dots, 1$}
    \State $\hat{\by}_t \leftarrow \by_t$; \; $\cK_{\text{spec}} \leftarrow \emptyset$
    \For{$j = 1, \dots, K_{\text{draft}}$}
        \State \textbf{Phase 0 (Draft):} $\{d_{t,j}^k\} \leftarrow p_{\btheta}^{\text{hard}}(\cdot \mid \bx, \hat{\by}_t;\, \cC_{\text{aux}})$ \Comment{cheap draft microstep with frozen/heuristic fast-path cache}
        \State $\bu_{t,j} \sim \pi_{\bphi}^{\text{unmask}}(\cdot \mid \hat{\by}_t)$ \Comment{initially this head can be frozen}
        \For{$k : u_{t,j}^k {=} 1,\, \hat{y}_t^k {=} \mask$}
            \State $\hat{y}_t^k \sim d_{t,j}^k$; \; $\cK_{\text{spec}} \leftarrow \cK_{\text{spec}} \cup \{(k, \hat{y}_t^k)\}$
        \EndFor
    \EndFor
    \State \textbf{Phase 0b (Validate):} $\{p_t^k\}, \{\rho_t^k\} \leftarrow p_{\btheta}^{\text{soft}}(\cdot \mid \bx, \hat{\by}_t),\, h_\psi(\hat{\by}_t)$ \Comment{one verifier pass over the drafted state}
    \State Compute $\alpha_t^k = \ind[\rho_t^k \ge \delta_{\text{acc}}]$; construct $s_t$ via \Cref{eq:softmask}
    \State $(\mathbf{r}_t, \bm{\kappa}_t, \mathbf{z}_t) \sim \pi_{\bphi}(\cdot \mid s_t)$; build $\cQ_t$ via \Cref{eq:Qt_candidates}
    \If{$\frac{1}{L}\sum_k (1 - \alpha_t^k) > \eta$}
        \State $\cC_{\text{aux}} \leftarrow \emptyset$ \Comment{cache realignment when the fast path becomes stale}
    \EndIf
    \For{$k : r_t^k {=} 1,\, \hat{y}_t^k \neq \mask$} \Comment{\textbf{Phase 1: Remask + evict from stable cache}}
        \State $\hat{y}_t^k \leftarrow \mask$; \; $\cK_{\text{stable}} \leftarrow \cK_{\text{stable}} \setminus \{k\}$
    \EndFor
    \State $\tilde{p}_t^k \leftarrow \textsc{Compose}(p_t^k, d_{t,K_{\text{draft}}}^k, \rho_t^k, \gamma)$ \Comment{\Cref{eq:composed}}
    \For{$k : \alpha_t^k {=} 0,\, r_t^k {=} 0$} \Comment{\textbf{Phase 2: Verifier correction / soft merge}}
        \State $\hat{y}_t^k \sim \tilde{p}_t^k$
    \EndFor
    \For{$k : \kappa_t^k {=} 1,\, r_t^k {=} 0$} \Comment{\textbf{Phase 3b: Stable cache (persistent, multi-step)}}
        \State $\cK_{\text{stable}} \leftarrow \cK_{\text{stable}} \cup \{k\}$
    \EndFor
    \State $\by_{t-1} \leftarrow \hat{\by}_t$
\EndFor
\State \textbf{Return} $\by_1$
\end{algorithmic}
\end{algorithm}

At each verifier event (\Cref{alg:aoae}), the auxiliary takes $K_{\text{draft}}$ cheap microsteps under a plug-and-play cache policy and accumulates a transient frontier $\cK_{\text{spec}}$ of drafted candidates.
The verifier then scores the drafted state once, accepts easy positions, remasks clear failures, softly corrects ambiguous positions, and may clear the drafter cache if the rejection rate is too high.
The method is deliberately lossy: wall-time savings come from running the cheaper cached drafter more often than the verifier, not from exact multi-step verifier simulation.

\paragraph{Two-pool cache semantics.}
Speculative acceptance and stable caching are different signals.
$\cK_{\text{spec}}$ contains candidates touched since the last verifier event and is replaced every verifier call.
$\cK_{\text{stable}}$ contains positions whose KVs the $\kappa_t$ head predicts will remain accurate over future steps, so it persists until explicit remasking.
Conflating the two would train $\kappa_t$ on a corrupted target, so AOAE maintains:
\begin{itemize}[itemsep=2pt]
    \item $\cK_{\text{spec}}$: transient speculative frontier; populated by freshly drafted candidates; replaced rather than accumulated.
    \item $\cK_{\text{stable}}$: persistent stable cache; populated by $\kappa_t^k = 1$ and evicted by explicit remask $r_t^k = 1$.
\end{itemize}
In the current $K_{\text{draft}}{=}1$ implementation, $\cK_{\text{spec}}$ can be materialized as a one-step accepted/reusable mask, but skipped positions are never allowed to refresh themselves: only positions freshly recomputed by the primary verifier can populate the next $\cK_{\text{spec}}$ frontier.
For $K_{\text{draft}}{>}1$, $\cK_{\text{spec}}$ should likewise be interpreted only as a verifier frontier, not a persistent cache.
Accordingly, the evaluation reports stable-cache occupancy, speculative-frontier occupancy, and their union separately from the stable-cache commit survival rate.
During initial experimentation, the unmasking and remasking heads can be frozen while cache/access heads are trained.
\paragraph{Prefix and drafter cache reuse.}
In practice, the prompt tokens $\bx$ are static for the entire generation.
We prefill the verifier's prompt KV cache once, storing KV vectors for all $P + L_{\text{gen}}$ positions.
The drafter also carries its own fast-path cache state $\cC_{\text{aux}}$, which may freeze or reuse KVs more aggressively according to any chosen heuristic or learned rule.
The verifier can clear or refresh $\cC_{\text{aux}}$ when the verifier-side rejection rate spikes, making speculative validation a natural cache-realignment mechanism.

Following~\cite{jazbec2025learning}, if $\bu_t = \mathbf{0}$ and $\cM_t \neq \emptyset$, the highest-confidence masked position is unmasked.
This fallback is disabled during training to prevent reward hacking.

\subsection{Composed Prediction}
\label{sec:composed}

The auxiliary and primary share weights and differ only in routing, so at low $\tau_r$ their predictions are highly correlated.
We compose the auxiliary's draft distribution $d_t^k$ with the verifier's $p_t^k$:
\begin{equation}
    \tilde{p}_t^k(v) \propto p_t^k(v) \cdot d_t^k(v)^{\gamma \cdot \rho_t^k},
    \label{eq:composed}
\end{equation}
where $\gamma \geq 0$ controls composition strength and $\rho_t^k \in [0,1]$ is the verifier-side keep score.
When the verifier strongly endorses the draft ($\rho_t^k \approx 1$), the composed distribution concentrates mass on the shared high-probability token, preserving the drafter's speed advantage.
When the verifier is uncertain or rejecting ($\rho_t^k \approx 0$), the distribution reduces to the verifier alone.
Since composition sharpens only where the verifier is supportive, the any-order property is preserved while still allowing soft correction rather than only hard accept/reject.

During GRPO training, composed prediction is active, so the policy learns to account for verifier acceptance and correction risk.
The cache-thrashing penalty naturally teaches the optional $\kappa_t$ head to cache only positions whose verifier-side scores remain stable.

\subsection{Policy Architecture}
\label{sec:architecture}

The steering policy $\pi_{\bphi}$ is a lightweight transformer:
\begin{enumerate}[leftmargin=*,itemsep=2pt]
    \item \textbf{Input:} Per-position features $(\bh_t^k, m_t^k, \rho_t^k, \alpha_t^k, g_t^k, \ell_t^k, t/T)$ projected to dimension $d_\pi$, where $g_t^k$ is a per-position age counter (steps since last action) and $\ell_t^k$ encodes the last action type.
    \item \textbf{Backbone:} $N_\pi$-layer transformer with $d_\pi$-dimensional hidden states and full bidirectional attention.
    \item \textbf{Heads:} Four linear projections producing logits for unmask ($f_{\bphi^{\text{unmask}}}$), remask ($f_{\bphi^{\text{remask}}}$), stable cache ($f_{\bphi^{\text{cache}}}$), and access/refresh ($f_{\bphi^{\text{access}}}$) per position. The verifier-side score supplies features and safety thresholds; the policy still chooses the stochastic actions.
\end{enumerate}
With $N_\pi = 1$ and $d_\pi = 128$, this is ${\sim}$500K parameters ($<$0.01\% of the base model).
The verifier-side score $\rho_t^k$ and its thresholded version $\alpha_t^k$ directly inform both unmask and cache logits.
Age features and last-action encodings allow the access head to model temporal revisit patterns caused by remasking and cache resets.

\subsection{Positional Speculative Caching}
\label{sec:positional_spec_cache}

The $\kappa_t$ head predicts \emph{which positions are stable enough to keep cached}; the access head $z_t^k$ predicts \emph{which positions will need compute soon}.
This second question is useful because diffusion decoding revisits positions in an instance-dependent pattern, especially when remasking is enabled.
Let $\cI := \{1,\dots,L\}$ be response positions and let $\cQ_t \subseteq \cI$ denote positions selected for refresh/prefetch.
Any position modified at step $t$ must be refreshed:
\begin{equation}
    \cQ_t \supseteq \{k : u_t^k = 1\} \cup \{k : r_t^k = 1\}.
    \label{eq:Qt_mandatory}
\end{equation}
Beyond these mandatory updates, a top-$B$ set of likely future accesses is refreshed speculatively:
\begin{equation}
    \cC_t^{(H)} := \text{Top-}B\bigl(\{g_{\bphi}^{(H)}(s_t)^k\}_{k=1}^L\bigr),
    \qquad
    \cQ_t := \cC_t^{(H)} \cup \{k : u_t^k = 1 \lor r_t^k = 1\},
    \label{eq:Qt_candidates}
\end{equation}
Unlike fixed windows or block schedules, $\cQ_t$ is predicted from $s_t$ rather than prescribed by position.
Thus AOAE treats diffusion caching as an access-pattern prediction problem: allocate refresh budget to positions likely to change soon without forcing the unmask/remask trajectory to become left-to-right or blockwise.
This mechanism is optional and model-agnostic; it can be used with any MDLM that exposes token-wise uncertainty and supports KV reuse.

We also optionally predict a layer boundary $\ell_t^\star$ for depth-selective refresh:
\begin{equation}
    \ell_t^\star \sim \text{Cat}\!\left(\text{softmax}(h_{\bphi}(s_t))\right),
    \qquad
    \text{recompute layers } \ell > \ell_t^\star \text{ for } k \in \cQ_t,
    \label{eq:learned_boundary_layer}
\end{equation}
Shallow boundaries save compute but risk stale deeper states; deeper boundaries are safer but slower.
We treat this as an ablation rather than a requirement for the core speculative method.

\subsection{Training via GRPO}
\label{sec:optimization}

The base model and PRISM adapter remain frozen; only the policy and soft-mask gating parameters are trained via GRPO~\cite{shao2024deepseekmath}.

\paragraph{Reward.}
The reward must credit both fewer verifier calls and within-step savings from caching.
The important bookkeeping point is that $\cK_{\text{spec}}$ and $\cK_{\text{stable}}$ are not interchangeable: $\cK_{\text{spec}}$ is a transient agreement frontier, while $\cK_{\text{stable}}$ is the learned persistent cache.
We therefore do not reward raw $\cK_{\text{spec}}$ occupancy as free compute.
Let $\bar{\kappa}_{\mathrm{op}}$ denote the mean \emph{operational} cache fraction used by the active implementation (zero in the conservative verifier-first setting, or the stable-cache fraction when stable KV reuse is enabled), and let:
\begin{equation}
    \text{eff\_flops} = \frac{\hat{T}}{T}\!\left(1 - \bar{\kappa}_{\mathrm{op}}\right),
    \label{eq:mean_cached}
\end{equation}
where $\hat{T}$ is the number of verifier events used.
The reward is:
\begin{equation}
    R = r(\by^*, \hat{\by}) \cdot (1 - \text{eff\_flops})^\alpha
      - \beta\, \overline{\text{Thrash}}
      - w_u\, f_{\text{unresolved}}
      + w_{F1}\, \overline{F1}_{\text{cache}}
      + w_q\, F1_{\text{access}},
    \label{eq:reward}
\end{equation}
where $r(\cdot)$ is task correctness and:

\begin{itemize}[itemsep=2pt]
    \item $\overline{\text{Thrash}}=\frac{1}{L}\sum_{t=1}^{T}\sum_k \ind[k \in \cK_{\text{stable},t-1} \land r_t^k=1]$ penalizes remasking stable-cached positions while keeping the term on the same scale as the bounded speed and correctness factors.
    \item $f_{\text{unresolved}}$ penalizes leftover $\mask$ tokens.
    \item $\overline{F1}_{\text{cache}}$ is a dense stability signal for $\kappa_t$: positions with low relative drift $\|\bH_t^k-\bH_{t-1}^k\|_2/\|\bH_{t-1}^k\|_2$ should be cached, and unstable ones should not.
    \item $F1_{\text{access}}$ measures whether $\cQ_t$ anticipates actual position modifications over the next $H$ steps.
\end{itemize}
The multiplicative structure of the speed term ensures incorrect generations receive zero reward regardless of speed, preventing the reward hacking observed with additive formulations~\cite{jazbec2025learning}.

\paragraph{GRPO objective.}
For each prompt, $G$ trajectories are sampled under $\pi_{\bphi_{\text{old}}}$ with greedy base-model decoding, so variation arises solely from the policy's stochastic actions.
With advantage $A^g = R^g - \bar{R}$ and importance ratio $\rho_t^g = \pi_{\bphi}(\mathbf{a}_t^g \mid s_t^g) / \pi_{\bphi_{\text{old}}}(\mathbf{a}_t^g \mid s_t^g)$:
\begin{equation}
    \cJ(\bphi) = \E_{\bx, \by^*} \left[ \frac{1}{G} \sum_{g=1}^G \frac{1}{T - \hat{T}_g} \sum_{t=\hat{T}_g}^{T} \min\!\Big\{ \rho_t^g A^g,\; \text{clip}(\rho_t^g, 1{-}\epsilon, 1{+}\epsilon)\, A^g \Big\} \right].
    \label{eq:grpo}
\end{equation}
KL regularization is omitted since the policy is trained from scratch.
In practice, we clamp log-ratios before exponentiation and optionally normalize advantages by within-group standard deviation to avoid numerical blowups at larger group sizes.
Behaviorally, this trains the policy to trust the drafter where the verifier agrees, defer to the verifier where it disagrees, and cache only positions whose verifier-side state remains stable.

% ======================================================================
\section{Experimental Design}
\label{sec:experiments}
% ======================================================================

\paragraph{Setup.}
LLaDA~2.1-mini (16B MoE, frozen) in dual-model configuration.
Policy: 1-layer transformer (${\sim}$500K parameters, input dimension $D{+}5$ including age and last-action features).
PRISM adapter fine-tuned on 10K samples.
Hyperparameters: $K{=}5$ (soft-masking), $G{=}8$, $\alpha{=}1.0$, $\beta{=}0.1$, $\epsilon{=}0.2$, $\gamma{=}0.5$, $w_{F1}{=}0.02$, $w_u{=}0.25$, and initially $w_q{=}0$ until the stable-cache head converges.
We sweep routing/verifier settings, verifier cadence, agreement threshold, unmask budget, remasking, drafter cache aggressiveness, draft horizon $K_{\text{draft}}$, and policy temperature $\tau_\pi$.
All evaluations run through the dInfer \cite{dinfer} integration when possible.

\paragraph{Datasets.}
GSM8K~\cite{cobbe2021gsm8k} and MATH~\cite{hendrycks2021math} for mathematical reasoning; HumanEval~\cite{chen2021humaneval} for code generation.
Policy training uses 50K prompts from OpenMathInstruct-2.

\paragraph{Metrics.}
Accuracy (exact match / pass@1), throughput (tokens/sec), NFEs, verifier agreement $\bar{\alpha}$, stable-cache occupancy, speculative-frontier occupancy, combined cache occupancy, stable-cache commit survival, drafter-cache reset rate, cache invalidation/thrashing, effective speedup vs.\ verifier-only decoding, and refresh ratio $\frac{1}{T}\sum_t |\cQ_t|/L$.

\paragraph{Ablations.}
We use four ablation groups.
\begin{itemize}[leftmargin=*,itemsep=1pt]
    \item \textbf{Speculation:} verifier-only; drafter-only; full draft-token speculation; location-only steering with primary tokens; $\gamma=0$ conservative composition.
    \item \textbf{Caching:} $\cK_{\text{spec}}$ only, $\cK_{\text{stable}}$ only, combined, learned vs.\ heuristic drafter cache, and cache-reset thresholds.
    \item \textbf{Policy:} no learned policy, no agreement feature, no remask, no access head, and $N_\pi \in \{1,2,4\}$.
    \item \textbf{Drafter efficiency:} hard-gated normal mode, S-mode, dKV/Elastic-style caching, and optional cheaper attention patterns such as sliding-window, sparse/strided, or linear attention.
\end{itemize}

% ======================================================================

\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{austin2021structured}
J.~Austin, D.~Johnson, J.~Ho, D.~Tarlow, and R.~van~den~Berg.
\newblock Structured denoising diffusion models in discrete state-spaces.
\newblock In \emph{NeurIPS}, 2021.

\bibitem{sahoo2024simple}
S.~Sahoo, M.~Arriola, Y.~Schiff, A.~Gokaslan, E.~Marber, S.~Kuleshov, C.~Jones, and V.~Kuleshov.
\newblock Simple and effective masked diffusion language models.
\newblock In \emph{NeurIPS}, 2024.

\bibitem{shi2024simplified}
J.~Shi, K.~Han, Z.~Wang, A.~Doucet, and M.~Titsias.
\newblock Simplified and generalized masked diffusion for discrete data.
\newblock In \emph{NeurIPS}, 2024.

\bibitem{lou2024discrete}
A.~Lou, C.~Meng, and S.~Ermon.
\newblock Discrete diffusion modeling by estimating the ratios of the data distribution.
\newblock In \emph{ICML}, 2024.

\bibitem{ou2024your}
J.~Ou, Y.~Song, and P.~Li.
\newblock Your absorbing discrete diffusion secretly models the conditional distributions of clean data.
\newblock \emph{arXiv:2406.03736}, 2024.

\bibitem{nie2025llada2}
S.~Nie, F.~Zhu, Z.~You, X.~Zhang, J.~Feng, Z.~Zhang, C.~Chen, G.~Xu, Y.~Dong, B.~Liu, et~al.
\newblock Large language diffusion models.
\newblock \emph{arXiv:2502.09992}, 2025.

\bibitem{llada21}
S.~Nie, F.~Zhu, C.~Chen, et~al.
\newblock {LLaDA 2.1}: Speeding up text diffusion via token editing.
\newblock \emph{arXiv:2602.08676}, 2025.

\bibitem{wu2025fastdllm}
R.~Wu, Y.~Qiu, D.~Zheng, Z.~Yu, L.~Jiang, and L.~Jiao.
\newblock Fast-{dLLM}: Training-free fast generation for diffusion large language models.
\newblock \emph{arXiv preprint}, 2025.

\bibitem{jazbec2025learning}
M.~Jazbec, T.~X.~Olausson, E.~Hu, P.~Hartout, D.~Yogatama, E.~Mathieu, and M.~Cuturi.
\newblock Learning unmasking policies for diffusion language models.
\newblock \emph{arXiv:2512.09106}, 2025.

\bibitem{kim2025trainworst}
J.~Kim, K.~Shah, V.~Kontonis, S.~Kakade, and S.~Chen.
\newblock Train for the worst, plan for the best: Understanding token ordering in masked diffusions.
\newblock In \emph{ICML}, 2025.

\bibitem{kim2025prism}
J.~Kim, J.~Li, B.~Chen, and T.~Hashimoto.
\newblock {PRISM}: Plug-in remasking for inference-time self-correction of masked diffusions.
\newblock \emph{arXiv:2510.01384}, 2025.

\bibitem{ma2025dkv}
Y.~Ma, C.~Li, Y.~Liu, and G.~Neubig.
\newblock {dKV-Cache}: The cache for diffusion language models.
\newblock In \emph{NeurIPS}, 2025.

\bibitem{elastic_cache}
Q.~Nguyen-Tri, M.~Ranjan, and Z.~Shen.
\newblock {Attention Is All You Need for KV Cache in Diffusion LLMs}.
\newblock In \emph{ICLR}, 2026.

\bibitem{hersche2025softmasked}
M.~Hersche, S.~Moor-Smith, T.~Hofmann, and A.~Rahimi.
\newblock Soft-masked diffusion language models.
\newblock \emph{arXiv:2510.17206}, 2025.

\bibitem{shao2024deepseekmath}
Z.~Shao, P.~Wang, Q.~Zhu, R.~Xu, J.~Song, M.~Zhang, Y.~K.~Li, Y.~Wu, and D.~Guo.
\newblock {DeepSeekMath}: Pushing the limits of mathematical reasoning in open language models.
\newblock \emph{arXiv:2402.03300}, 2024.

\bibitem{cobbe2021gsm8k}
K.~Cobbe, V.~Kosaraju, M.~Bavarian, M.~Chen, H.~Jun, L.~Kaiser, M.~Plappert, J.~Tworek, J.~Hilton, R.~Nakano, C.~Hesse, and J.~Schulman.
\newblock Training verifiers to solve math word problems.
\newblock \emph{arXiv:2110.14168}, 2021.

\bibitem{hendrycks2021math}
D.~Hendrycks, C.~Burns, S.~Kadavath, A.~Arora, S.~Basart, E.~Tang, D.~Song, and J.~Steinhardt.
\newblock Measuring mathematical problem solving with the {MATH} dataset.
\newblock In \emph{NeurIPS}, 2021.

\bibitem{chen2021humaneval}
M.~Chen, J.~Tworek, H.~Jun, Q.~Yuan, H.~Pinto, J.~Kaplan, H.~Edwards, Y.~Burda, N.~Joseph, G.~Brockman, et~al.
\newblock Evaluating large language models trained on code.
\newblock \emph{arXiv:2107.03374}, 2021.

\bibitem{dinfer}
Y.~Ma, L.~Du, L.~Wei, K.~Chen, Q.~Xu, K.~Wang, et~al.
\newblock dInfer: An Efficient Inference Framework for Diffusion Language Models.
\newblock \emph{arXiv:2510.08666}, 2025.

\end{thebibliography}
\end{document}
